Effect of the Nominal Value of Foreign Currencies in valuing a good

### 2024-04-16 paper.tex
first draft

---
title: "Effect of the Nominal Value of Foreign Currencies in valuing a good"
date: 2024-04-16
author:
  - name: Konrad Kopp
    email: konrad.kopp@stcatz.ox.ac.uk
    affiliation: 
      - name: University of Oxford
abstract: > 
  Research into the effect of using foreign currencies on consumption behaviour suggests that consumers systematically under- or overspend based on the nominal (face) value of the foreign currency. However, previous research is undecided on which direction this effect has, with the two most prominent models leading to different and mutually exclusive outcomes. This paper considers the different models and tests their predictive and explanatory accuracy using a study. It finds that individuals tend to underspend when the foreign currency is less numerous than (a fraction of) the home currency and overspend when the opposite is the case. 
license: "CC0"
copyright: 
  holder: Konrad Kopp
  year: 2024
keywords:
  - Behavioural Economics
  - Face Value Effect
  - Consumer Choice Theory
---

# Introduction

Past experience suggests that we treat foreign currencies differently to our home currency and have difficulty adjusting to them. Previous research, such as the "money illusion" effect suggests that individuals are biased towards the nominal (or face value) of prices. Past research has investigated whether a systemic effect

RAS [@Raghubir2002] examined the effect of using foreign currencies on consumption behaviour and argue that there is a systematic difference in spending behaviour based on the exchange rate of the foreign currency to an individuals' home currency. Specifically, they showed that when a foreign currency is a fraction of the home currency (e.g. 1 unit in home currency is 0.2 units in foreign currency) consumers tend to overspend in real terms relative to their home currency. On the other hand, when the foreign currency is a multiple of the home currency (e.g. 1 unit in home currency is 5 units in foreign currency) consumers tend to underspend in real terms.^[For consistency, I will use the terms fraction and multiple currencies throughout this paper as defined here.] RAS argue that this effect arises because individuals are anchored on the face value of the foreign currency which influences the exchange rate calculations from and to their home currencies. Hence, RAS term the result of this inadequate adjustment the face value effect.

WER [@Wertenbroch2007] consider the findings by RAS and reiterate their conclusion that individuals inadequately calculate the values of goods between differnet currencies, leading to systematic differences in spending behaviour based on the exchange rate between the foreign and home currencies. However, WER argue that the process of adjustment occurs differently, namely that individuals calculate their valuations of goods based on salient reference points, such as budget constraints. Contrary to standard theory, WER argue that consumers care about the difference between a valuation and their budget rather than the ratio of the two. In the context of foreign currencies, the nominal value of this difference will diverge, leading to systematic differences in spending behaviour. However, the direction of this bias is the exact opposite of the one shown by RAS. Specifically, WER show that consumers underspend when the foreign currency is a fraction of the home currency and overspend in the opposite case.

Fang [@Fang2019] further builds on top of the research by RAS and aims to extend it to the domain of "virtual" currencies, which refers to non-fiat currencies that are digital in nature and not controlled and issued by a central bank. Examples of virtual currencies include video game (or in-game) currencies, rewards points, airmiles, cryptocurrencies and more to the extent that these are at least somewhat liquid. Investigating in-game currencies specifically, Fang shows that consumer valuations systematically deviate when using these currencies relative to the home currency. Fang's findings correspond with those of RAS and the face value effect.

This paper aims to build on the aforementioned research in several important ways. Foremost, previous research is conclusive that there is an effect that using foreign currencies has on individuals' consumption behaviour relative to using their home currencies, but is divided on the direction of this effect. Hence, this paper attempts to evaluate the different models posed and use a study to test which model most adequately predicts the perceived behaviour. Second, previous research has only assigned minor significance to using incentive-compatible mechanisms to elicit true preferences of participants. These mechanisms have been established to elicit more accurate responses from participants that are closer to their true preferences [@cite]. Hence, this study will use the Becker-DeGroot-Marshak (BDM) mechanism to elicit participants' willingness to pay (WTP).[^ The BDM mechanism will be defined and explained in more detail in the Methodology section below.] RAS [@Raghubir2002] expressed the concern that "despite attempts to make the tasks as realistic as possible, the studies reported were laboratory experiments, and issues of generalizability when real money is on the line do arise. For instance, when people actually exchange foreign currency they may feel richer or poorer depending on the exchange rate, and the differences in perceptions of wealth may affect product valuation". Similarly, Fang [@Fang2019] noted that implementing the BDM mechanism was infeasible for the study he conducted but that future research should focus on using something like it to elicit WTP. Finally, this paper aims to extend the previous literature on the effects of using virtual currencies specifically, due to their increasing rise in importance when transacting or making purchasing decisions online.

@@

# Theoretical Framework

## Standard Model

Standard consumer choice theory suggests that individuals formulate valuations of goods in relation to reference points, such as their budget [@cite]. Specifically, standard theory poses that consumers evaluate their valuation for a good as a ratio with the budget in order to determine what proportion of their spending power is required to purchase the good. To formaulate valuations of goods in a different currency, consumers would first determine the value of the good in their home currency and subsequently apply the exchange rate to get their valuation in the foreign currency. Hence, the model follows a two step process:

1. Formaulate valuation of good as ratio to budget in home currency
2. Convert the valuation into foreign currency

@@

$$V_n = f(\dfrac{x}{M_n})$$
$$V = V_n * X = V_r$$

where V is the valuation in the foreign currency, $V_n$ is the valuation of the good in the home currency (nominal), $V_r$ is the valuation of the good in the foreign currency, $M_n$ is the budget in the home currency, X is the exchange rate and $f(x)$ is some function that uses a valuation relative to the budget in order to compute $V_n$.

@@

## Face Value Effect Model

RAS [@Raghubir2002] put forth a different model for how consumers formulate their valuation of a good in a foreign currency based on their studies conducted. Specifically, they pose that a consumer will first formulate a valuation of a good in their home currency, before converting that amount into the foreign currency using the exchange rate. However, they argue that this calculation is biased by the nominal value of the good in the home currency (in cases where a consumer does not determine their valuation but observes a price, this calculation would be biased by the nominal value of the good in the foreign currency). This model may be formulated as follows:

1. Formulate valuation of good in home currency
2. Convert the valuation into foreign currency
3. Adjust using the nominal value of good in home currency

@@

$$V = α(V_n) + (1 - α)V_r$$

where V is the valuation in the foreign currency, $V_n$ is the valuation of the good in the home currency (nominal), $V_r$ is the valuation of the good in the foreign currency and α is a weighting parameter such that $α ∈[0,1]$. Importantly, RAS [@Raghubir2002] argue that $α > 0$ and that V is hence not equal to $V_r$ but is biased towards $V_n$.

## Perceived Value of Money Model

WER [@Wertenbroch2007] build on top of RAS, but put forth a different model that they argue is more representative of the behaviour of actual consumers. Specifically, they agree with the standard model that valuations are made in relation to salient reference points, such as a consumers' budget, and argue that the model put forth by RAS fails to account for this. However, they disagree with the standard model that valuations use a ratio of price to budget, but instead argue that consumers think in terms of differences. Hence, instead of considering what fraction of their purchasing power is required to buy a good, consumers consider how much of their budget is left over. However, WER argue that consumers inadequately adjust this difference to the exchange rate and are thus biased when considering goods in different currencies since the nominal differences between price and budget will differ when considering different currencies. This model can be formulated as follows:

1. Formaulate valuation of good as difference to budget in home currency
2. Convert the valuation into foreign currency
3. Adjust using the difference to budget in the foreign currency

@@

$$V = α(M_r - V_r) + (1 - α)V_r$$

where V is the valuation in the foreign currency, $V_n$ is the valuation of the good in the home currency (nominal), $V_r$ is the valuation of the good in the foreign currency, $M_r$ is the budget in the foreign currency (real) and α is a weighting parameter such that $α ∈[0,1]$. Importantly, WER [@Wertenbroch2007] argue that $α > 0$ and that V is hence not equal to $V_r$ but is biased towards $M_r - V_r$.

## Predictions

@@

- Standard model: $V_r = 5 (125) BlueCoin$
- Face value effect model: $V_r > 5 (< 125) BlueCoin$
- Perceived value of money model: $V_r < 5 (> 125) BlueCoin$

where $V_r$ is the valuation of the good in the foreign currency.

@@

- Standard model: $V_rf = V_rm$
- Face value effect model: $V_rf > V_rm$
- Perceived value of money model: $V_rf < V_rm$

where $V_rf$ is the valuation of the good in the foreign currency for the fraction group and $V_rm$ is the valuation of the good in the foreign currency for the multiple group.

@@

# Method

@@

## Study

@@

The auction was the main part of the study, with the aim of eliciting the participants' WTP on a specific item and comparing these across different currencies used. In this section, participants were randomly split into two groups with almost-identical instructions. The participants were instructed to bid on the item shown below, a tabletop airhockey table, using the virtual currency "BlueCoin". The auction was explained to follow the BDM mechanism and it was also stated that three participants of the experiment would be randomly selected, their auction would be simulated and they would receive the outcome. The only difference between the groups was the exchange rate between BlueCoin and the British Pound (£) and the endowment that the participants received in BlueCoin. The first group received an endowment of 10 BlueCoin where 1 BlueCoin equals £5 and the second group received an endowment of 250 BlueCoin where 1 BlueCoin equals £0.2. To complete the auction, participants were asked to respond with their bid, in BlueCoin.

@@

## Auction Design

Since the aim of the study was to determine whether there were any systematic differences in WTP between consumers when the only differing factor is the currency used, we decided to not include a control group that bid on the item using Pounds, which was the design of some of the experiments run by RAS [@Raghubir2002]. This allowed for a simpler design and more power given the same sample size. Further, unlike the studies run by RAS [@Raghubir2002] and WER [@Wertenbroch2007], we did not use a fiat currency or mix of fiat and fictional currencies, but used only one fictional currency for both groups. On the one hand, this simplification was due to a focus of wether the effects described in previous papers also held in similar ways for virtual currencies. On the other hand, it also allowed us to preclude any biases or previous experience that participants might have had towards or with certain currencies. RAS [@Raghubir2002] acknowledge that "the face value effect is due to the accessibility and perceptual salience of the face value of the foreign currency ... [which] is likely to depend on the extent to which an individual has the opportunity or the time available to process exchange rate information and/or has experience in using a particular foreign currency". To bias participants the least, we chose the relatively nondescript name for a fictional, virtual currency: BlueCoin.

@@

In designing the main part of the experiment, we chose to conduct an auction, specifically using the BDM mechanism, in order to have an incentive-compatible method of eliciting the participants' WTP. The BDM's incentive compatibility in eliciting WTP is a well-established phenomenon that is superior to using participants' stated preferences [@cite]. As mentioned above, the studies conducted by RAS [@Raghubir2002], WER [@Wertenbroch2007] (with the exception of one permutation of one of the studies) and Fang [@Fang2019] did not implement incentive-compatible methods of eliciting the participants' preferences, making this experiment and important addition to these previous studies in examining whether the effects describe still hold when a more accurate way of eliciting WTP is used. In order to incentivise the participants even further, we ran a lottery that paid out three randomly-chosen participants based on their actual choices made in the auction and by simulating the BDM mechanism using a random number generator. Due to budget constraints, we were only able to play this lottery for three people, but given that the sample size of the experiment was relatively small, the potential to receive one's outcome is likely to have helped in eliciting the participants' true WTP.

Finally, given our budget, we were able to give participants an endowment of £50 each, which was around double of what the item costs to buy. Assuming that the item is priced roughly around the average WTP (which turned out to be correct), this endowment was chosen to leave enough room for people with higher WTP to make their bid and for us to capture this in the data. This is important for the goals of the experiment, since if we had, for example, chosen an endowment of £20, the majority of participants might have bid their maximum amount, hindering our ability to meaningfully compare WTP across the two groups. We chose relatively simple exchange rates of 5 and $1/5$ so that the exchange rate calculations would be relatively easy to make, allowing us to be more confident that differences between groups were not due to the mental effort required to do calculations but due to something else. RAS [@Raghubir2002] acknowledge that "the reliance on face value may be a function of the ease with which the foreign money can be converted. ... Future research should examine the asymmetric and nonlinear nature of this effect". Since the goal of this study is to investigate that an effect exists that is not solely based on a high difficulty or large mental effort required when doing specific, difficult calculations, we aimed to make the calculations as easy as possible in order to isolate the effect of the exchange rate and nominal value of the foreign currency.

## Limitations

In setting up the experiment, we incorrectly used the BDM mechanism. Instead of explaining it as a sealed-bid second-price auction, the experiment instructions explained it as a sealed-bid first-price auction. Unlike in a second-price auction, the participants of a first price auction do not have a dominant strategy to bid their true valuations, in fact there generally is no dominant strategy in such an auction [@cite]. [@todo]

@@

# Results

@@

```{r}
# Load the experiment data
experiment_data <-
    read.csv("data/experiment_data_formatted.csv")

# Summary statistics for the bid columns
summary(experiment_data$group_fraction_bid_in_pounds)
sd(experiment_data$group_fraction_bid_in_pounds, na.rm = TRUE)

summary(experiment_data$group_multiple_bid_in_pounds)
sd(experiment_data$group_multiple_bid_in_pounds, na.rm = TRUE)
```

@@

```{r}
# Load the experiment data
experiment_data <-
    read.csv("data/experiment_data_formatted.csv")

# Plot a boxplot for the bid columns
boxplot(
    experiment_data$group_fraction_bid_in_pounds,
    experiment_data$group_multiple_bid_in_pounds,
    main = "Boxplot of fraction and multiple group bids (in Pounds)",
    names = c("Group Fraction", "Group Multiple"),
    xlab = "Variables",
    ylab = "Values"
)
```

@@

$$h_0: µ_1 = µ_2$$
$$h_1: µ_1 ≠ µ_2$$

@@

# References
::: {#refs}
:::

# Appendix {.appendix}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{report}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[lmargin=30mm,rmargin=30mm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{british}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Effect of the Nominal Value of Foreign Currencies in valuing a good},
  pdfauthor={Konrad Kopp},
  pdflang={en-GB},
  pdfkeywords={Behavioural Economics, Face Value Effect, Consumer Choice
Theory},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Effect of the Nominal Value of Foreign Currencies in valuing a
good}

  \author{Konrad Kopp}
            \affil{%
                  University of Oxford
              }

\date{2024-04-16}
\begin{document}
\maketitle
\begin{abstract}
Research into the effect of using foreign currencies on consumption
behaviour suggests that consumers systematically under- or overspend
based on the nominal (face) value of the foreign currency. However,
previous research is undecided on which direction this effect has, with
the two most prominent models leading to different and mutually
exclusive outcomes. This paper considers the different models and tests
their predictive and explanatory accuracy using a study. It finds that
individuals tend to underspend when the foreign currency is less
numerous than (a fraction of) the home currency and overspend when the
opposite is the case.
\end{abstract}

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{2}
\tableofcontents
}
\chapter{Introduction}\label{introduction}

Past experience suggests that we treat foreign currencies differently to
our home currency and have difficulty adjusting to them. Previous
research, such as the ``money illusion'' effect suggests that
individuals are biased towards the nominal (or face value) of prices.
Past research has investigated whether a systemic effect

RAS (Raghubir and Srivastava 2002) examined the effect of using foreign
currencies on consumption behaviour and argue that there is a systematic
difference in spending behaviour based on the exchange rate of the
foreign currency to an individuals' home currency. Specifically, they
showed that when a foreign currency is a fraction of the home currency
(e.g.~1 unit in home currency is 0.2 units in foreign currency)
consumers tend to overspend in real terms relative to their home
currency. On the other hand, when the foreign currency is a multiple of
the home currency (e.g.~1 unit in home currency is 5 units in foreign
currency) consumers tend to underspend in real terms.\footnote{For
  consistency, I will use the terms fraction and multiple currencies
  throughout this paper as defined here.} RAS argue that this effect
arises because individuals are anchored on the face value of the foreign
currency which influences the exchange rate calculations from and to
their home currencies. Hence, RAS term the result of this inadequate
adjustment the face value effect.

WER (Wertenbroch, Soman, and Chattopadhyay 2007) consider the findings
by RAS and reiterate their conclusion that individuals inadequately
calculate the values of goods between differnet currencies, leading to
systematic differences in spending behaviour based on the exchange rate
between the foreign and home currencies. However, WER argue that the
process of adjustment occurs differently, namely that individuals
calculate their valuations of goods based on salient reference points,
such as budget constraints. Contrary to standard theory, WER argue that
consumers care about the difference between a valuation and their budget
rather than the ratio of the two. In the context of foreign currencies,
the nominal value of this difference will diverge, leading to systematic
differences in spending behaviour. However, the direction of this bias
is the exact opposite of the one shown by RAS. Specifically, WER show
that consumers underspend when the foreign currency is a fraction of the
home currency and overspend in the opposite case.

Fang (Fang 2019) further builds on top of the research by RAS and aims
to extend it to the domain of ``virtual'' currencies, which refers to
non-fiat currencies that are digital in nature and not controlled and
issued by a central bank. Examples of virtual currencies include video
game (or in-game) currencies, rewards points, airmiles, cryptocurrencies
and more to the extent that these are at least somewhat liquid.
Investigating in-game currencies specifically, Fang shows that consumer
valuations systematically deviate when using these currencies relative
to the home currency. Fang's findings correspond with those of RAS and
the face value effect.

This paper aims to build on the aforementioned research in several
important ways. Foremost, previous research is conclusive that there is
an effect that using foreign currencies has on individuals' consumption
behaviour relative to using their home currencies, but is divided on the
direction of this effect. Hence, this paper attempts to evaluate the
different models posed and use a study to test which model most
adequately predicts the perceived behaviour. Second, previous research
has only assigned minor significance to using incentive-compatible
mechanisms to elicit true preferences of participants. These mechanisms
have been established to elicit more accurate responses from
participants that are closer to their true preferences (\textbf{cite?}).
Hence, this study will use the Becker-DeGroot-Marshak (BDM) mechanism to
elicit participants' willingness to pay (WTP).{[}\^{} The BDM mechanism
will be defined and explained in more detail in the Methodology section
below.{]} RAS (Raghubir and Srivastava 2002) expressed the concern that
``despite attempts to make the tasks as realistic as possible, the
studies reported were laboratory experiments, and issues of
generalizability when real money is on the line do arise. For instance,
when people actually exchange foreign currency they may feel richer or
poorer depending on the exchange rate, and the differences in
perceptions of wealth may affect product valuation''. Similarly, Fang
(Fang 2019) noted that implementing the BDM mechanism was infeasible for
the study he conducted but that future research should focus on using
something like it to elicit WTP. Finally, this paper aims to extend the
previous literature on the effects of using virtual currencies
specifically, due to their increasing rise in importance when
transacting or making purchasing decisions online.

@@

\chapter{Theoretical Framework}\label{theoretical-framework}

\section{Standard Model}\label{standard-model}

Standard consumer choice theory suggests that individuals formulate
valuations of goods in relation to reference points, such as their
budget (\textbf{cite?}). Specifically, standard theory poses that
consumers evaluate their valuation for a good as a ratio with the budget
in order to determine what proportion of their spending power is
required to purchase the good. To formaulate valuations of goods in a
different currency, consumers would first determine the value of the
good in their home currency and subsequently apply the exchange rate to
get their valuation in the foreign currency. Hence, the model follows a
two step process:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Formaulate valuation of good as ratio to budget in home currency
\item
  Convert the valuation into foreign currency
\end{enumerate}

@@

\[V_n = f(\dfrac{x}{M_n})\] \[V = V_n * X = V_r\]

where V is the valuation in the foreign currency, \(V_n\) is the
valuation of the good in the home currency (nominal), \(V_r\) is the
valuation of the good in the foreign currency, \(M_n\) is the budget in
the home currency, X is the exchange rate and \(f(x)\) is some function
that uses a valuation relative to the budget in order to compute
\(V_n\).

@@

\section{Face Value Effect Model}\label{face-value-effect-model}

RAS (Raghubir and Srivastava 2002) put forth a different model for how
consumers formulate their valuation of a good in a foreign currency
based on their studies conducted. Specifically, they pose that a
consumer will first formulate a valuation of a good in their home
currency, before converting that amount into the foreign currency using
the exchange rate. However, they argue that this calculation is biased
by the nominal value of the good in the home currency (in cases where a
consumer does not determine their valuation but observes a price, this
calculation would be biased by the nominal value of the good in the
foreign currency). This model may be formulated as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Formulate valuation of good in home currency
\item
  Convert the valuation into foreign currency
\item
  Adjust using the nominal value of good in home currency
\end{enumerate}

@@

\[V = α(V_n) + (1 - α)V_r\]

where V is the valuation in the foreign currency, \(V_n\) is the
valuation of the good in the home currency (nominal), \(V_r\) is the
valuation of the good in the foreign currency and α is a weighting
parameter such that \(α ∈[0,1]\). Importantly, RAS (Raghubir and
Srivastava 2002) argue that \(α > 0\) and that V is hence not equal to
\(V_r\) but is biased towards \(V_n\).

\section{Perceived Value of Money
Model}\label{perceived-value-of-money-model}

WER (Wertenbroch, Soman, and Chattopadhyay 2007) build on top of RAS,
but put forth a different model that they argue is more representative
of the behaviour of actual consumers. Specifically, they agree with the
standard model that valuations are made in relation to salient reference
points, such as a consumers' budget, and argue that the model put forth
by RAS fails to account for this. However, they disagree with the
standard model that valuations use a ratio of price to budget, but
instead argue that consumers think in terms of differences. Hence,
instead of considering what fraction of their purchasing power is
required to buy a good, consumers consider how much of their budget is
left over. However, WER argue that consumers inadequately adjust this
difference to the exchange rate and are thus biased when considering
goods in different currencies since the nominal differences between
price and budget will differ when considering different currencies. This
model can be formulated as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Formaulate valuation of good as difference to budget in home currency
\item
  Convert the valuation into foreign currency
\item
  Adjust using the difference to budget in the foreign currency
\end{enumerate}

@@

\[V = α(M_r - V_r) + (1 - α)V_r\]

where V is the valuation in the foreign currency, \(V_n\) is the
valuation of the good in the home currency (nominal), \(V_r\) is the
valuation of the good in the foreign currency, \(M_r\) is the budget in
the foreign currency (real) and α is a weighting parameter such that
\(α ∈[0,1]\). Importantly, WER (Wertenbroch, Soman, and Chattopadhyay
2007) argue that \(α > 0\) and that V is hence not equal to \(V_r\) but
is biased towards \(M_r - V_r\).

\section{Predictions}\label{predictions}

@@

\begin{itemize}
\tightlist
\item
  Standard model: \(V_r = 5 (125) BlueCoin\)
\item
  Face value effect model: \(V_r > 5 (< 125) BlueCoin\)
\item
  Perceived value of money model: \(V_r < 5 (> 125) BlueCoin\)
\end{itemize}

where \(V_r\) is the valuation of the good in the foreign currency.

@@

\begin{itemize}
\tightlist
\item
  Standard model: \(V_rf = V_rm\)
\item
  Face value effect model: \(V_rf > V_rm\)
\item
  Perceived value of money model: \(V_rf < V_rm\)
\end{itemize}

where \(V_rf\) is the valuation of the good in the foreign currency for
the fraction group and \(V_rm\) is the valuation of the good in the
foreign currency for the multiple group.

@@

\chapter{Method}\label{method}

@@

\section{Study}\label{study}

@@

The auction was the main part of the study, with the aim of eliciting
the participants' WTP on a specific item and comparing these across
different currencies used. In this section, participants were randomly
split into two groups with almost-identical instructions. The
participants were instructed to bid on the item shown below, a tabletop
airhockey table, using the virtual currency ``BlueCoin''. The auction
was explained to follow the BDM mechanism and it was also stated that
three participants of the experiment would be randomly selected, their
auction would be simulated and they would receive the outcome. The only
difference between the groups was the exchange rate between BlueCoin and
the British Pound (£) and the endowment that the participants received
in BlueCoin. The first group received an endowment of 10 BlueCoin where
1 BlueCoin equals £5 and the second group received an endowment of 250
BlueCoin where 1 BlueCoin equals £0.2. To complete the auction,
participants were asked to respond with their bid, in BlueCoin.

@@

\section{Auction Design}\label{auction-design}

Since the aim of the study was to determine whether there were any
systematic differences in WTP between consumers when the only differing
factor is the currency used, we decided to not include a control group
that bid on the item using Pounds, which was the design of some of the
experiments run by RAS (Raghubir and Srivastava 2002). This allowed for
a simpler design and more power given the same sample size. Further,
unlike the studies run by RAS (Raghubir and Srivastava 2002) and WER
(Wertenbroch, Soman, and Chattopadhyay 2007), we did not use a fiat
currency or mix of fiat and fictional currencies, but used only one
fictional currency for both groups. On the one hand, this simplification
was due to a focus of wether the effects described in previous papers
also held in similar ways for virtual currencies. On the other hand, it
also allowed us to preclude any biases or previous experience that
participants might have had towards or with certain currencies. RAS
(Raghubir and Srivastava 2002) acknowledge that ``the face value effect
is due to the accessibility and perceptual salience of the face value of
the foreign currency \ldots{} {[}which{]} is likely to depend on the
extent to which an individual has the opportunity or the time available
to process exchange rate information and/or has experience in using a
particular foreign currency''. To bias participants the least, we chose
the relatively nondescript name for a fictional, virtual currency:
BlueCoin.

@@

In designing the main part of the experiment, we chose to conduct an
auction, specifically using the BDM mechanism, in order to have an
incentive-compatible method of eliciting the participants' WTP. The
BDM's incentive compatibility in eliciting WTP is a well-established
phenomenon that is superior to using participants' stated preferences
(\textbf{cite?}). As mentioned above, the studies conducted by RAS
(Raghubir and Srivastava 2002), WER (Wertenbroch, Soman, and
Chattopadhyay 2007) (with the exception of one permutation of one of the
studies) and Fang (Fang 2019) did not implement incentive-compatible
methods of eliciting the participants' preferences, making this
experiment and important addition to these previous studies in examining
whether the effects describe still hold when a more accurate way of
eliciting WTP is used. In order to incentivise the participants even
further, we ran a lottery that paid out three randomly-chosen
participants based on their actual choices made in the auction and by
simulating the BDM mechanism using a random number generator. Due to
budget constraints, we were only able to play this lottery for three
people, but given that the sample size of the experiment was relatively
small, the potential to receive one's outcome is likely to have helped
in eliciting the participants' true WTP.

Finally, given our budget, we were able to give participants an
endowment of £50 each, which was around double of what the item costs to
buy. Assuming that the item is priced roughly around the average WTP
(which turned out to be correct), this endowment was chosen to leave
enough room for people with higher WTP to make their bid and for us to
capture this in the data. This is important for the goals of the
experiment, since if we had, for example, chosen an endowment of £20,
the majority of participants might have bid their maximum amount,
hindering our ability to meaningfully compare WTP across the two groups.
We chose relatively simple exchange rates of 5 and \(1/5\) so that the
exchange rate calculations would be relatively easy to make, allowing us
to be more confident that differences between groups were not due to the
mental effort required to do calculations but due to something else. RAS
(Raghubir and Srivastava 2002) acknowledge that ``the reliance on face
value may be a function of the ease with which the foreign money can be
converted. \ldots{} Future research should examine the asymmetric and
nonlinear nature of this effect''. Since the goal of this study is to
investigate that an effect exists that is not solely based on a high
difficulty or large mental effort required when doing specific,
difficult calculations, we aimed to make the calculations as easy as
possible in order to isolate the effect of the exchange rate and nominal
value of the foreign currency.

\section{Limitations}\label{limitations}

In setting up the experiment, we incorrectly used the BDM mechanism.
Instead of explaining it as a sealed-bid second-price auction, the
experiment instructions explained it as a sealed-bid first-price
auction. Unlike in a second-price auction, the participants of a first
price auction do not have a dominant strategy to bid their true
valuations, in fact there generally is no dominant strategy in such an
auction (\textbf{cite?}). (\textbf{todo?})

@@

\chapter{Results}\label{results}

@@

\begin{verbatim}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
   0.00   10.00   20.00   19.74   32.00   50.00      20 
\end{verbatim}

\begin{verbatim}
[1] 15.49001
\end{verbatim}

\begin{verbatim}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
   0.00   16.50   27.50   25.57   35.25   50.00      21 
\end{verbatim}

\begin{verbatim}
[1] 14.8083
\end{verbatim}

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-2-1.pdf}

@@

\[h_0: µ_1 = µ_2\] \[h_1: µ_1 ≠ µ_2\]

@@

\chapter{References}\label{references}
